---
title: "Walkability Index"
authors: Cheyna Warner, Erika Davis, & Kyla Brannigan
output:
  flexdashboard::flex_dashboard:
    theme:
      version: 4
      bg: "#FFF5F5"
      fg: "#000000" 
      navbar-bg: "#FFCFE5"
      base_font: 
        google: Quicksand
      heading_font:
        google: Quicksand
      code_font:
        google: 
          family: Quicksand
          local: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rio)
library(flexdashboard)
library(shiny)
library(tidyverse)
library(here)
library(ggplot2)
library(ggrepel) 

```

```{r, echo=FALSE}
# # Walkability Data Cleaning
# smart_location <- import(here("EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv")) #smart location database and walkability index.
# 
# walkability <- smart_location %>%
#   select(GEOID10, CBSA, CBSA_Name, CBSA_POP, CBSA_EMP, CBSA_WRK, # select col of interest
#          TotPop, AutoOwn0, AutoOwn1, AutoOwn2p,
#          NatWalkInd,
#          TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Svc, E5_Ent,
#          E_LowWageWk, E_MedWageWk, E_HiWageWk)
# 
# OR_walkability <- walkability |>
#   filter(str_detect(CBSA_Name, " OR")) |>
#   arrange(CBSA_Name)
# write.csv(OR_walkability, "OR_walkability.csv")
# 
# capitals_list <- c("Montgomery, AL", "Juneua, AK", "Phoenix, AZ", "Little Rock, AK", "Sacramento, CA", "Denver, CO", "Hartford, CT", "Dover, DE", "Tallahassee, FL", "Atlanta, GA", "Honolulu, HI", "Boise, ID", "Springfield, MO", "Indianapolis, IN", "Des Moines, ID", "Topeka, KS", "Frankfort, KY", "Baton Rouge, LA", "Annapolis, MD", "Boston, MA", "Lansing, MI", "Saint Paul, MN", "Jackson, MS", "Jefferson City, MO", "Helena, MT", "Lincon, NE", "Carson City, NV", "Concord, NH", "Trenton, NJ", "Santa Fe, NM", "Albany, NY", "Raleigh, NC", "Bismarck, ND", "Coloumbus, OH", "Oklahoma City, OK", "Salem, OR", "Harrisburg, PA", "Providence, RI", "Columbia, SC", "Pierre, SD", "Nashville, TN", "Austin, TX","Salt Lake City, UT", "Montpelier, VT", "Richmond, VA", "Olympia, WA", "Charleston, WV", "Madison, WI", "Cheyenne, WY")
# 
# walkability <- walkability %>% # only selecting data for state capitals
#   filter(CBSA_Name %in% capitals_list)
# 
# walkability$City <- gsub(",.*$", "", walkability$CBSA_Name) # make new col that indicates city
# walkability$State <- sub("^.*?, \\s*", "", walkability$CBSA_Name) # make new col that indicates state
# 
# write.csv(walkability, "Walkability.csv")
```

```{r, echo=FALSE}
crash <- import(here("data", "FARS2019NationalCSV", "accident.csv")) #crash data

walkability <- import(here("data", "Walkability.csv")) #cleaned walkability data

OR_walkability<- import(here("data", "OR_walkability.csv")) #Oregon-specific walkability

city_walkability <- walkability %>% #walkability for each city
  group_by(City) %>% 
  summarise(mean_walk = mean(NatWalkInd))

# Crash Data Filtering
crash <- crash %>% 
  select(STATENAME, RUR_URBNAME, HARM_EVNAME, FATALS)
```

Lifestyle and Walkability
======
completed by Kyla Brannigan and Cheyna Warner

```{r, echo=FALSE}
### Initial plan: Map with color gradient showing concentration of walkable cities
# I can't get this to work right now so I'm pivoting a bit, but I will try again later (I really want to display a map)
# # library-ing packages needed for mapping
# library(tigris)
# library(sf)
```

```{r, echo=FALSE}
# # Load walkability data
# walk <- OR_walkability
#   mutate(GEOID10 = formatC(as.numeric(GEOID10), format = "fg", flag = "0") |> 
#            str_pad(width = 12, side = "left", pad = "0"))
# 
# # Get Oregon block group geometries
# or_bg <- block_groups(state = "OR", year = 2019, cb = TRUE)
# 
# # Join walkability data to block group geometries
# or_bg_walk <- or_bg |> 
#   left_join(walk, by = c("GEOID" = "GEOID10"))
# 
# # Plot
# ggplot(or_bg_walk) +
#   geom_sf(aes(fill = NatWalkInd), color = NA) +
#   scale_fill_viridis_c(
#     name = "Walkability\nIndex",
#     option = "plasma",
#     na.value = "grey80"
#   ) +
#   labs(
#     title = "Oregon Walkability by Block Group",
#     subtitle = "National Walkability Index (EPA Smart Location Database)",
#     caption = "Source: EPA Smart Location Database"
#   ) +
#   theme_void() +
#   theme(legend.position = "right")
```

### How does Worker Wage Concentration and Walkability 

```{r, echo=FALSE}
# looking at concentration of low, med, and high wage workers and walkability

# adding the pre-defined categories for walkability to make it more interpretable rather than the continuous scale
wage_walkability <- walkability |> 
  filter(!is.na(NatWalkInd), !is.na(E_LowWageWk), !is.na(E_MedWageWk), !is.na(E_HiWageWk)) |> 
  mutate(WalkCat = cut(NatWalkInd,
    breaks = c(0, 5.75, 10.5, 15.25, 20),
    labels = c("Least\nWalkable\n(1–5.75)", "Below Avg\n(5.75–10.5)", "Above Avg\n(10.5–15.25)", "Most\nWalkable\n(15.25–20)"), 
    include.lowest = TRUE
  ))

# summarize by walkability bin (doing median instead of mean to help it not be skewed by places like NYC or rural areas)
summary_wage_walkability <- wage_walkability |> 
  group_by(WalkCat) |> 
  summarise(
    Low  = median(E_LowWageWk, na.rm = TRUE),
    Med  = median(E_MedWageWk, na.rm = TRUE),
    High = median(E_HiWageWk,  na.rm = TRUE),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = c(Low, Med, High),
               names_to = "WageGroup",
               values_to = "MedianWorkers") |> 
  mutate(WageGroup = factor(WageGroup, levels = c("Low", "Med", "High")))
```

```{r, echo=FALSE}
wage_walkability_plot <- ggplot(summary_wage_walkability, aes(x = WalkCat, y = MedianWorkers, fill = WageGroup)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.92) +
  labs(title    = "Worker Wage Concentration by Neighborhood Walkability",
       subtitle = "Median number of workers per census block group across walkability quartiles") +
  theme_minimal(base_size = 13) 

wage_walkability_plot
```

### do more walkable cities have fewer cars? 

```{r, echo=FALSE}
car_rates <- walkability %>% 
  select(CBSA, CBSA_Name, AutoOwn0, AutoOwn1, AutoOwn2p,NatWalkInd, City, State) %>% 
  group_by(CBSA_Name) %>% 
  mutate(car = sum(AutoOwn1+AutoOwn2p),
         mean_walk = mean(NatWalkInd))

car_rates <- car_rates[!duplicated(car_rates$CBSA_Name), ]

car_rates_plot1 <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm")
car_rates_plot1

car_rates_plot2 <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")

car_rates_plot2

car_rates_plot3 <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Relationship between walkability and number of households that own over 1 car",
       x= "Walkability",
       y = "Number of Households wiith > 1 car")

car_rates_plot3

car_rates_plot_final <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Relationship between walkability and number of households that own over 1 car",
       x= "Walkability",
       y = "Number of Households wiith > 1 car")

car_rates_plot_final
```

### are unemployment rates lower in walkable areas?

```{r, echo=FALSE}
employment_rates <- walkability %>% 
  select(CBSA, CBSA_Name, TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Ind, E5_Svc, E5_Ent, State, City, NatWalkInd) %>% 
  group_by(CBSA_Name) %>% 
  mutate(mean_walk = mean(NatWalkInd))

employment_rates <- employment_rates[!duplicated(employment_rates$CBSA_Name), ]

employment_rates_plot1 <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm")
  
employment_rates_plot1

employment_rates_plot2 <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")

employment_rates_plot2

employment_rates_plot3 <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Relationship between walkability and employment rates",
       x= "Walkability",
       y = "Employment Rate")

employment_rates_plot3

employment_rates_plot <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Relationship between walkability and employment rates",
       x= "Walkability",
       y = "Employment Rate")

employment_rates_plot
```

Pedestrian crash data and Walkability 
====== 
completed by Erika Davis

Column {data-width=650}
----------------------------
### What is the relationship between walkability and pedestrian crash data? 

```{r, echo=FALSE}

#adding a state column to the city_walkability as the ped. crashes only has states

state_vector <- c("Louisana", "North Dakota", "Nevada", "West Virgina", "Wyoming", "South Carolina", "New Hampshire", "Delaware", "Kentucky", "Montana", "Mississippi", "Missouri", "Wisconsin", "Alabama", "Oklahoma", "South Dakota", "Virginia", "Oregon", "Utah", "New Mexico", "Missouri", "Florida", "Kansas")

state_walkability <- city_walkability %>%
  mutate(state = state_vector)

#summarizing fatalities by state
state_fatalities <- crash %>%
  group_by(STATENAME) %>%
  summarize(total_fatals = sum(FATALS, na.rm = TRUE))

crash_df <- state_walkability %>%
  left_join(state_fatalities, by = c("state" = "STATENAME"))

#defining outliers so i can label them in the plot
crash_df <- crash_df %>%
  mutate(label_state = case_when(
    total_fatals > quantile(total_fatals, 0.95, na.rm = TRUE) ~ state,       #top 5% fatalities
    mean_walk > quantile(mean_walk, 0.95, na.rm = TRUE) ~ state, #top 5% walkability
    mean_walk < quantile(mean_walk, 0.05, na.rm = TRUE) ~ state, #bottom 5% walkability
    TRUE ~ NA_character_                                                     
  ))

#plot
ggplot(crash_df, aes(x = mean_walk, y = total_fatals)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  geom_text_repel(aes(label = label_state), 
                  box.padding = 0.5, 
                  point.padding = 0.3, 
                  segment.color = 'grey50') +
  geom_smooth(method = "lm", color = "pink", linetype = "dashed", se = FALSE) +
  labs(
    title = "Walkability vs. Pedestrian Fatalities by State",
    subtitle = "Labels shown for statistical extremes (top/bottom 5%)",
    x = "Walkability Score",
    y = "Total Fatalities"
  ) +
  theme_minimal()

```

Column {data-width=350}
----------------------------

### Do rural or urban places have more traffic fatalities? 

```{r, echo=FALSE}

crash_filtered <- crash %>% 
  filter(RUR_URBNAME %in% c("Rural", "Urban")) 
  
ggplot(crash_filtered, aes(x = RUR_URBNAME, y = FATALS, fill = RUR_URBNAME)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values = c("Rural" = "forestgreen", "Urban" = "pink")) +
  labs(
    title = "Total Fatalities: Rural vs. Urban",
    x = "Location Type",
    y = "Total Fatalities",
    fill = "Legend"
  ) +
  theme_minimal()
```

Column {data-width=350}
----------------------------
### Do rural or urban states have more traffic fatalities? 

```{r, echo=FALSE}
ggplot(crash_filtered, aes(x = RUR_URBNAME, y = FATALS, fill = RUR_URBNAME)) +
  geom_bar(stat = "summary", fun = "sum") +
  # This line creates the separate panels for each state
  facet_wrap(~ `STATENAME`) + 
  scale_fill_manual(values = c("Rural" = "forestgreen", "Urban" = "pink")) +
  labs(
    title = "Fatalities by State: Rural vs. Urban",
    x = "Location Type",
    y = "Total Fatalities",
    fill = "Legend"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

Column {}
----------------------------
