---
title: "Walkability Index"
authors: Cheyna Warner, Erika Davis, & Kyla Brannigan
output:
  flexdashboard::flex_dashboard:
    theme:
      version: 4
      bg: "#FFF5F5"
      fg: "#000000" 
      primary: "#5E9C07"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rio)
library(tidyverse)
library(here)
```

# Importing and Clearning the Data

Here, we are importing and cleaning our data for analysis.
```{r}
smart_location <- import(here("data", "EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv")) #smart location database and walkability index 


crash <- import(here("FARS2019NationalCSV", "accident.csv")) #crash data

#select measures of interest from smart_location and crash data
walkability <- smart_location %>% 
  select(CBSA, CBSA_Name, CBSA_POP, CBSA_EMP, CBSA_WRK, 
         TotPop, AutoOwn0, AutoOwn1, AutoOwn2p, 
         NatWalkInd, 
         TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Svc, E5_Ent,
         E_LowWageWk, E_MedWageWk, E_HiWageWk)

crash <- crash %>% 
  select(STATENAME, RUR_URBNAME, HARM_EVNAME, FATALS)
```

# Location and Walkability
completed by Kyla Brannigan

### Map with color gradient showing concentration of walkable cities

```{r}

```


# Pedestrian crash data and Walkability 
completed by Erika Davis

###  What is the relationship between walkability and pedestrian crash data? 

```{r}

```

### How are crash rates impacted by the most common mode of transportation? (bar graph)

```{r}

```
 

# Lifestyle and Walkability
completed by Cheyna Warner

### do more walkable cities have fewer cars? 

```{r}
car_rates <- walkability %>% 
  select(CBSA, CBSA_Name, AutoOwn0, AutoOwn1, AutoOwn2p,NatWalkInd) %>% 
  group_by(CBSA, NatWalkInd) %>% 
  mutate(car = sum(AutoOwn1+AutoOwn2p))

car_rates$car <- as.factor(car_rates$car)

car_rates_plot <- car_rates %>% 
  ggplot(aes(x = car, y = NatWalkInd))+
  geom_point()

car_rates_plot
```
 
### are unemployment rates lower in walkable areas?

```{r}

```

