---
title: "Walkability Index"
authors: Cheyna Warner, Erika Davis, & Kyla Brannigan
output: 
  flexdashboard::flex_dashboard:
    theme:
      version: 4
      bg: "#FFF5F5"
      fg: "#000000" 
      primary: "#5E9C07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rio)
library(tidyverse)
library(here)
```

# Importing and Clearning the Data
This chunk was used initially to clean the raw data, but for deployment the data file is too large to put on github and must be unzipped to run as a webpage. 
```{r}
# # Walkability Data Cleaning
# smart_location <- import(here("EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv")) #smart location database and walkability index.
# 
# walkability <- smart_location %>%
#   select(GEOID10, CBSA, CBSA_Name, CBSA_POP, CBSA_EMP, CBSA_WRK, # select col of interest
#          TotPop, AutoOwn0, AutoOwn1, AutoOwn2p,
#          NatWalkInd,
#          TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Svc, E5_Ent,
#          E_LowWageWk, E_MedWageWk, E_HiWageWk)
# 
# OR_walkability <- walkability |>
#   filter(str_detect(CBSA_Name, " OR")) |>
#   arrange(CBSA_Name)
# write.csv(OR_walkability, "OR_walkability.csv")
# 
# capitals_list <- c("Montgomery, AL", "Juneua, AK", "Phoenix, AZ", "Little Rock, AK", "Sacramento, CA", "Denver, CO", "Hartford, CT", "Dover, DE", "Tallahassee, FL", "Atlanta, GA", "Honolulu, HI", "Boise, ID", "Springfield, MO", "Indianapolis, IN", "Des Moines, ID", "Topeka, KS", "Frankfort, KY", "Baton Rouge, LA", "Annapolis, MD", "Boston, MA", "Lansing, MI", "Saint Paul, MN", "Jackson, MS", "Jefferson City, MO", "Helena, MT", "Lincon, NE", "Carson City, NV", "Concord, NH", "Trenton, NJ", "Santa Fe, NM", "Albany, NY", "Raleigh, NC", "Bismarck, ND", "Coloumbus, OH", "Oklahoma City, OK", "Salem, OR", "Harrisburg, PA", "Providence, RI", "Columbia, SC", "Pierre, SD", "Nashville, TN", "Austin, TX","Salt Lake City, UT", "Montpelier, VT", "Richmond, VA", "Olympia, WA", "Charleston, WV", "Madison, WI", "Cheyenne, WY")
# 
# walkability <- walkability %>% # only selecting data for state capitals
#   filter(CBSA_Name %in% capitals_list)
# 
# walkability$City <- gsub(",.*$", "", walkability$CBSA_Name) # make new col that indicates city
# walkability$State <- sub("^.*?, \\s*", "", walkability$CBSA_Name) # make new col that indicates state
# 
# write.csv(walkability, "Walkability.csv")
```

Here, we are importing and cleaning our data for analysis.
```{r}
crash <- import(here("FARS2019NationalCSV", "accident.csv")) #crash data

walkability <- import(here("Walkability.csv")) #cleaned walkability data

OR_walkability<- import(here("OR_walkability.csv")) #Oregon-specific walkability

city_walkability <- walkability %>% #walkability for each city
  group_by(City) %>% 
  summarise(mean_walk = mean(NatWalkInd))

# Crash Data Filtering
crash <- crash %>% 
  select(STATENAME, RUR_URBNAME, HARM_EVNAME, FATALS)
```

# Location and Walkability
completed by Kyla Brannigan

### Map with color gradient showing concentration of walkable cities

```{r}
# library-ing packages needed for mapping
library(tigris)
library(sf)
```

```{r}
# get Oregon block group geometries from census
or_bg <- block_groups(state = "OR", year = 2019, cb = TRUE)

# aggregate walkability to county
walk_summary <- OR_walkability %>%
  group_by(CBSA, CBSA_Name) %>%
  summarise(avg_walkability = mean(NatWalkInd, na.rm = TRUE),
            total_pop = sum(TotPop, na.rm = TRUE))

# get CBSA boundaries
or_cbsa <- core_based_statistical_areas(year = 2019, cb = TRUE) %>%
  filter(GEOID %in% unique(OR_walkability$CBSA))

# join CBSA-level map
or_cbsa_walk <- or_cbsa %>%
  left_join(walk_summary, by = c("GEOID" = "CBSA"))

# base Oregon outline
or_state <- states(cb = TRUE) %>% filter(STUSPS == "OR")

ggplot() +
  geom_sf(data = or_state, fill = "grey90", color = "grey60") +
  geom_sf(data = or_cbsa_walk, aes(fill = avg_walkability), color = "white", size = 0.3) +
  scale_fill_viridis_c(name = "Avg. Walk\nIndex", option = "plasma", na.value = "grey80") +
  labs(
    title = "Oregon Walkability by Metro Area",
    subtitle = "National Walkability Index (EPA Smart Location Database)",
    caption = "Source: EPA Smart Location Database"
  ) +
  theme_void() +
  theme(legend.position = "right")
```


# Pedestrian crash data and Walkability 
completed by Erika Davis

###  What is the relationship between walkability and pedestrian crash data? 

```{r}

```

### How are crash rates impacted by the most common mode of transportation? (bar graph)

```{r}

```
 

# Lifestyle and Walkability
completed by Cheyna Warner

### do more walkable cities have fewer cars? 

```{r}
car_rates <- walkability %>% 
  select(CBSA, CBSA_Name, AutoOwn0, AutoOwn1, AutoOwn2p,NatWalkInd, City, State) %>% 
  group_by(CBSA_Name) %>% 
  mutate(car = sum(AutoOwn1+AutoOwn2p),
         mean_walk = mean(NatWalkInd))

car_rates <- car_rates[!duplicated(car_rates$CBSA_Name), ]

car_rates_plot <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Relationship between walkability and number of households that own over 1 car",
       x= "Walkability",
       y = "Number of Households wiith > 1 car")

car_rates_plot
```
 
### are unemployment rates lower in walkable areas?

```{r}
employment_rates <- walkability %>% 
  select(CBSA, CBSA_Name, TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Ind, E5_Svc, E5_Ent, State, City, NatWalkInd) %>% 
  group_by(CBSA_Name) %>% 
  mutate(mean_walk = mean(NatWalkInd))

employment_rates <- employment_rates[!duplicated(employment_rates$CBSA_Name), ]

employment_rates_plot <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Relationship between walkability and employment rates",
       x= "Walkability",
       y = "Employment Rate")

employment_rates_plot
```

