---
title: "Impact of Walkability"
authors: Cheyna Warner, Erika Davis, & Kyla Brannigan
output:
  flexdashboard::flex_dashboard:
    theme:
      version: 4
      bg: "#FFFFFF"
      fg: "#000000" 
      primary: "#82BDAB"
      navbar-bg: "#6bb599"
      base_font: 
        google: Quicksand
      heading_font:
        google: Quicksand
      code_font:
        google: 
          family: Quicksand
          local: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rio)
library(flexdashboard)
library(shiny)
library(tidyverse)
library(here)
library(ggplot2)
library(ggrepel) 
library(showtext)

font_add_google("Quicksand") # Add googoo Font
showtext_auto()

theme_set(theme_minimal(base_family = "Quicksand"))

options(scipen=999) #disable scientific notation 
```

```{r, echo=FALSE}
# # Walkability Data Cleaning
# smart_location <- import(here("EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv")) #smart location database and walkability index.
# 
# walkability <- smart_location %>%
#   select(GEOID10, CBSA, CBSA_Name, CBSA_POP, CBSA_EMP, CBSA_WRK, # select col of interest
#          TotPop, AutoOwn0, AutoOwn1, AutoOwn2p,
#          NatWalkInd,
#          TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Svc, E5_Ent,
#          E_LowWageWk, E_MedWageWk, E_HiWageWk)
# 
# OR_walkability <- walkability |>
#   filter(str_detect(CBSA_Name, " OR")) |>
#   arrange(CBSA_Name)
# write.csv(OR_walkability, "OR_walkability.csv")
# 
# capitals_list <- c("Montgomery, AL", "Juneua, AK", "Phoenix, AZ", "Little Rock, AK", "Sacramento, CA", "Denver, CO", "Hartford, CT", "Dover, DE", "Tallahassee, FL", "Atlanta, GA", "Honolulu, HI", "Boise, ID", "Springfield, MO", "Indianapolis, IN", "Des Moines, ID", "Topeka, KS", "Frankfort, KY", "Baton Rouge, LA", "Annapolis, MD", "Boston, MA", "Lansing, MI", "Saint Paul, MN", "Jackson, MS", "Jefferson City, MO", "Helena, MT", "Lincon, NE", "Carson City, NV", "Concord, NH", "Trenton, NJ", "Santa Fe, NM", "Albany, NY", "Raleigh, NC", "Bismarck, ND", "Coloumbus, OH", "Oklahoma City, OK", "Salem, OR", "Harrisburg, PA", "Providence, RI", "Columbia, SC", "Pierre, SD", "Nashville, TN", "Austin, TX","Salt Lake City, UT", "Montpelier, VT", "Richmond, VA", "Olympia, WA", "Charleston, WV", "Madison, WI", "Cheyenne, WY")
# 
# walkability <- walkability %>% # only selecting data for state capitals
#   filter(CBSA_Name %in% capitals_list)
# 
# walkability$City <- gsub(",.*$", "", walkability$CBSA_Name) # make new col that indicates city
# walkability$State <- sub("^.*?, \\s*", "", walkability$CBSA_Name) # make new col that indicates state
# 
# write.csv(walkability, "Walkability.csv")
```

```{r, echo=FALSE}
crash <- import(here("data", "FARS2019NationalCSV", "accident.csv")) #crash data

walkability <- import(here("data", "Walkability.csv")) #cleaned walkability data

OR_walkability<- import(here("data", "OR_walkability.csv")) #Oregon-specific walkability

city_walkability <- walkability %>% #walkability for each city
  group_by(City) %>% 
  summarise(mean_walk = mean(NatWalkInd))

# Crash Data Filtering
crash <- crash %>% 
  select(STATENAME, RUR_URBNAME, HARM_EVNAME, FATALS)
```

<style>
.sidebar h2 {
  font-size: 16px; /* Adjust size as needed */
}
.sidebar {
  font-size: 16px; /* Adjust size as needed */
}
</style>

Column {.sidebar}
==========================
**About Our Project**
---------------
This project combines 2019 data from the [Smart Location Database](https://catalog.data.gov/dataset/smart-location-database8) and [National Highway Safety Administration Pedestrian Crash Data](https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/FARS/) to provide a snapshot of how walkability influences multiple facets of life. We were most interested on seeing if city walkability is related to a number of things, including employment, average number of cars owned, average wages, and combining the Smart Location Data with the Pedestrian Crash data to see if there is a relationship between each. Links to each database can be found below. 

U.S. Environmental Protection Agency Smart Location Database (with Walkability Index)
------------------------

https://catalog.data.gov/dataset/smart-location-database8

National Highway Safety Administration Pedestrian Crash Data
--------------------------

https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/FARS/

Wage
==========================
Completed by Kyla Brannigan

Column {.tabset .tabset-fade}
-------------------------------------

```{r, echo=FALSE}
### Initial plan: Map with color gradient showing concentration of walkable cities
# I can't get this to work right now so I'm pivoting a bit, but I will try again later (I really want to display a map)
# # library-ing packages needed for mapping
# library(tigris)
# library(sf)
```

```{r, echo=FALSE}
# # Load walkability data
# walk <- OR_walkability
#   mutate(GEOID10 = formatC(as.numeric(GEOID10), format = "fg", flag = "0") |> 
#            str_pad(width = 12, side = "left", pad = "0"))
# 
# # Get Oregon block group geometries
# or_bg <- block_groups(state = "OR", year = 2019, cb = TRUE)
# 
# # Join walkability data to block group geometries
# or_bg_walk <- or_bg |> 
#   left_join(walk, by = c("GEOID" = "GEOID10"))
# 
# # Plot
# ggplot(or_bg_walk) +
#   geom_sf(aes(fill = NatWalkInd), color = NA) +
#   scale_fill_viridis_c(
#     name = "Walkability\nIndex",
#     option = "plasma",
#     na.value = "grey80"
#   ) +
#   labs(
#     title = "Oregon Walkability by Block Group",
#     subtitle = "National Walkability Index (EPA Smart Location Database)",
#     caption = "Source: EPA Smart Location Database"
#   ) +
#   theme_void() +
#   theme(legend.position = "right")
```

### Wage vs. Walkability

```{r, echo=FALSE}
# looking at concentration of low, med, and high wage workers and walkability

# adding the pre-defined categories for walkability to make it more interpretable rather than the continuous scale
wage_walkability <- walkability |> 
  filter(!is.na(NatWalkInd), !is.na(E_LowWageWk), !is.na(E_MedWageWk), !is.na(E_HiWageWk)) |> 
  mutate(WalkCat = cut(NatWalkInd,
    breaks = c(0, 5.75, 10.5, 15.25, 20),
    labels = c("Least\nWalkable\n(1–5.75)", "Below Avg\n(5.75–10.5)", "Above Avg\n(10.5–15.25)", "Most\nWalkable\n(15.25–20)"), 
    include.lowest = TRUE
  ))

# summarize by walkability bin (doing median instead of mean to help it not be skewed by places like NYC or rural areas)
summary_wage_walkability <- wage_walkability |> 
  group_by(WalkCat) |> 
  summarise(
    Low  = median(E_LowWageWk, na.rm = TRUE),
    Med  = median(E_MedWageWk, na.rm = TRUE),
    High = median(E_HiWageWk,  na.rm = TRUE),
    .groups = "drop"
  ) |> 
  pivot_longer(cols = c(Low, Med, High),
               names_to = "WageGroup",
               values_to = "MedianWorkers") |> 
  mutate(WageGroup = factor(WageGroup, levels = c("Low", "Med", "High")))
```

```{r, echo=FALSE}
wage_colors <- c("Low" = "#88CCEE", "Med" = "#DDCC77", "High" = "#CC6677") # color-blind friendly!

wage_walkability_plot <- summary_wage_walkability |> 
  ggplot(aes(x = WalkCat, y = MedianWorkers, fill = WageGroup)) +
  scale_fill_manual(values = wage_colors,
                    labels = c("Low", "Medium", "High"),
                    name = "Wage Group") +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.92) +
  labs(title    = "Worker Wage Concentration Differs by Neighborhood Walkability",
       subtitle = "Median number of workers per census block group across\nwalkability quartiles",
       x        = "National Walkability Index",
       y        = "Median Number of Workers\n(per census block)") +
  theme_minimal(base_size = 10) +
  theme(text = element_text(family = "Quicksand")) #make font macth the dashboard 

wage_walkability_plot
```

```{r echo=FALSE}
agg <- aggregate(cbind(E_LowWageWk, E_MedWageWk, E_HiWageWk) ~ WalkCat,
                 data = wage_walkability, FUN = sum, na.rm = TRUE)

cont_table <- as.matrix(agg[, c("E_LowWageWk", "E_MedWageWk", "E_HiWageWk")])
rownames(cont_table) <- agg$WalkCat
colnames(cont_table) <- c("Low Wage", "Med Wage", "High Wage")

# chi-square test
chisq_result <- chisq.test(cont_table)
#print(chisq_result)

# where are the biggest deviations
#print(round(chisq_result$stdres, 3))
```

Column {data-width=450 .tabset .tabset-fade}
-------------------------------------
### About 

This plot shows the relationship between walkability and worker wage concentration per [US Census Block](https://www.census.gov/newsroom/blogs/random-samplings/2011/07/what-are-census-blocks.html)

**Research Question:** How is walkability related to worker wage concentration? 

**Hypothesis:** I hypothesize that the distribution of worker wages will differ based on neighborhood walkability. 

**Conclusions:** Worker wage concentration distribution differs based on neighborhood walkability. A Chi-square test was conducted to determine this (X^2^ (6) = 24,210, *p* < .001). 

Specifically, the least walkable areas are skewed toward high-wage workers and underrepresent low-wage workers (though visually, this is hard to see). 

### v1

```{r, echo=FALSE}
wage_walkability_plot <- summary_wage_walkability |> 
  ggplot(aes(x = WalkCat, y = MedianWorkers, fill = WageGroup)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.92) +
  labs(title    = "Worker Wage Concentration by Neighborhood Walkability",
       subtitle = "Median number of workers per census block group across walkability quartiles") +
  theme(text = element_text(family = "Quicksand")) #make font macth the dashboard 

wage_walkability_plot
```

### v2
```{r, echo=FALSE}
wage_colors <- c("Low" = "#88CCEE", "Med" = "#DDCC77", "High" = "#CC6677") # color-blind friendly!

wage_walkability_plot <- summary_wage_walkability |> 
  ggplot(aes(x = WalkCat, y = MedianWorkers, fill = WageGroup)) +
  scale_fill_manual(values = wage_colors,
                    labels = c("Low", "Medium", "High"),
                    name = "Wage Group") +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.92) +
  labs(title    = "Worker Wage Concentration Differs by Neighborhood Walkability",
       subtitle = "Median number of workers per census block group across\nwalkability quartiles",
       x        = "National Walkability Index",
       y        = "Median Number of Workers\n(per census block)") +
  theme_minimal(base_size = 10) +
  theme(text = element_text(family = "Quicksand")) #make font macth the dashboard 

wage_walkability_plot
```

Employment
==========================
Completed by Cheyna Warner

Column {data-width=550 .tabset .tabset-fade}
-------------------------------------

```{r, echo=FALSE}
employment_rates <- walkability %>% 
  select(CBSA, CBSA_Name, TotEmp, E5_Ret, E5_Off, E5_Ind, E5_Ind, E5_Svc, E5_Ent, State, City, NatWalkInd) %>% 
  group_by(CBSA_Name) %>% 
  mutate(mean_walk = mean(NatWalkInd))

employment_rates <- employment_rates[!duplicated(employment_rates$CBSA_Name), ]
```

### How is walkability related to employment rates? 
```{r, echo=FALSE}

employment_rates_plot <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Walkable cities experience lower rates of employment",
       x= "Walkability",
       y = "Employment Rate")+
  theme(text = element_text(family = "Quicksand")) + #make font macth the dashboard 
  theme(legend.position="bottom")+
  theme(legend.key.size = unit(0.2, "cm")) 

employment_rates_plot
```

```{r, echo=FALSE}
p_car_rates <- cor.test(employment_rates$mean_walk, employment_rates$TotEmp, method = "pearson")$p.value
rho_car_rates <- cor.test(employment_rates$mean_walk, employment_rates$TotEmp, method = "pearson")$estimate
```

Column {data-width=450 .tabset .tabset-fade}
-------------------------------------
### About 

This plot shows the relationship between walkability and employment rates. Only data for capital cities is shown and only some capital cities were included in the dataset. 

**Research Question:** How is walkability related to employment rates? 

**Hypothesis:** I hypothesize that employment rates will be lower in less walkable areas. 

**Conclusions:** Walkability and employment rates have a significant, negative relationship. Therefore, more walkable areas have lower employment rates (aka higher unemployment rates).

### v1
```{r, echo=FALSE}
employment_rates_plot1 <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm")
  
employment_rates_plot1
```

### v2
```{r, echo=FALSE}
employment_rates_plot2 <- employment_rates %>% 
  ggplot(aes(x = mean_walk, y = TotEmp))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")

employment_rates_plot2
```


Cars 
====================
Completed by Cheyna Warner

Column {data-width=550 .tabset .tabset-fade}
-------------------------------------

```{r, echo=FALSE}
car_rates <- walkability %>% 
  select(CBSA, CBSA_Name, AutoOwn0, AutoOwn1, AutoOwn2p,NatWalkInd, City, State) %>% 
  group_by(CBSA_Name) %>% 
  mutate(car = sum(AutoOwn1+AutoOwn2p),
         mean_walk = mean(NatWalkInd))

car_rates <- car_rates[!duplicated(car_rates$CBSA_Name), ]
```

### Do households in walkable areas own fewer cars? 
```{r, echo=FALSE}
car_rates_plot_final <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")+
  theme_minimal()+
  labs(title = "Households have more cars in more walkable cities",
       x= "Walkability",
       y = "Number of Households with > 1 car")+
  theme(text = element_text(family = "Quicksand"))+
  theme(legend.position="bottom")+
  theme(legend.key.size = unit(0.2, "cm"))

car_rates_plot_final
```

```{r, echo=FALSE}
p_car_rates <- cor.test(car_rates$mean_walk, car_rates$car, method = "pearson")$p.value
rho_car_rates <- cor.test(car_rates$mean_walk, car_rates$car, method = "pearson")$estimate
```


Column {data-width=450 .tabset .tabset-fade}
-------------------------------------
### About

This graph shows the relationship between walkability and number of households that own at least 1 car. Only data for capital cities is shown and only some capital cities were included in the dataset. 

**Research Question:** Do households in walkable areas own fewer cars? 

**Hypothesis:** I hypothesize that people who live in walkable areas will have fewer cars. 

**Conclusions:** Walkability and number of cars have a significant, positive relationship. 


### v1

```{r, echo=FALSE}
car_rates_plot1 <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm")

car_rates_plot1
```

### v2

```{r, echo=FALSE}
car_rates_plot2 <- car_rates %>% 
  ggplot(aes(x = mean_walk, y = car))+
  geom_point(aes(color = City)) +
  geom_smooth(method ="lm", color = "black", fill = "pink")

car_rates_plot2
```

Walkability and Pedestrian Crashes
=====================================
Completed by Erika Davis

Column {data-width=550 .tabset .tabset-fade}
--------------------------------------
### What is the relationship between walkability and pedestrian crash data? 

```{r, echo=FALSE}

#adding a state column to the city_walkability as the ped. crashes only has states

state_vector <- c("Louisana", "North Dakota", "Nevada", "West Virgina", "Wyoming", "South Carolina", "New Hampshire", "Delaware", "Kentucky", "Montana", "Mississippi", "Missouri", "Wisconsin", "Alabama", "Oklahoma", "South Dakota", "Virginia", "Oregon", "Utah", "New Mexico", "Missouri", "Florida", "Kansas")

state_walkability <- city_walkability %>%
  mutate(state = state_vector)

#summarizing fatalities by state
state_fatalities <- crash %>%
  group_by(STATENAME) %>%
  summarize(total_fatals = sum(FATALS, na.rm = TRUE))

crash_df <- state_walkability %>%
  left_join(state_fatalities, by = c("state" = "STATENAME"))

#defining outliers so i can label them in the plot
crash_df <- crash_df %>%
  mutate(label_state = case_when(
    total_fatals > quantile(total_fatals, 0.95, na.rm = TRUE) ~ state,       #top 5% fatalities
    mean_walk > quantile(mean_walk, 0.95, na.rm = TRUE) ~ state, #top 5% walkability
    mean_walk < quantile(mean_walk, 0.05, na.rm = TRUE) ~ state, #bottom 5% walkability
    TRUE ~ NA_character_                                                     
  ))

#plot
ggplot(crash_df, aes(x = mean_walk, y = total_fatals)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.7) +
  geom_text_repel(aes(label = label_state), 
                  box.padding = 0.5, 
                  point.padding = 0.3, 
                  segment.color = 'grey50') +
  geom_smooth(method = "lm", color = "pink", linetype = "dashed", se = FALSE) +
  labs(
    title = "Walkability vs. Pedestrian Fatalities by State",
    subtitle = "Labels shown for statistical extremes (top/bottom 5%)",
    x = "Walkability Score",
    y = "Total Fatalities"
  ) +
  theme_minimal()

crashmodel <- lm(total_fatals ~ mean_walk, data = crash_df)


```

Column {data-width=450 .tabset .tabset-fade}
--------------------------------------

### About

This graph shows the relationship between how walkable cities are, and how many pedestrian fatalities happened in one year (2019). Data for walkability was originally by city, so was recoded to be by state, and not all states were represented in the dataset. According to the [EPA](https://www.epa.gov/smartgrowth/smart-location-mapping), this data was only available in places that share transit data in [GTFS format](https://developers.google.com/transit/gtfs/).

**Research Question:** Is there a relationship between walkability of states and the amount of pedestrian fatalities?

**Hypothesis:** I hypothesize that more walkable states will have less pedestrian fatalities.

**Conclusions:** There is no relationship between state walkability and the amount of pedestrian fatalities in one year. A simple linear regression was run to test this (*t*(19) = -0.17, *p* = 0.86).

### v1 
```{r, echo=FALSE}

ggplot(crash_df, aes(x = mean_walk, y = total_fatals)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm") + 
  theme_minimal()

```

### v2 
```{r, echo=FALSE}
 ggplot(crash_df, aes(x = mean_walk, y = total_fatals)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", linetype = "dashed", se = FALSE) +
  labs(
    title = "Walkability vs. Pedestrian Fatalities by State",
    subtitle = "Labels shown for statistical extremes (top/bottom 5%)",
    x = "Walkability Score",
    y = "Total Fatalities"
  ) +
  theme_minimal()
```

Rural vs. Urban Pedestrian Crashes
=====================================
Completed by Erika Davis

Column {data-width=550 .tabset .tabset-fade}
--------------------------------------

### Do rural or urban places have more traffic fatalities? 

```{r, echo=FALSE}

crash_filtered <- crash %>% 
  filter(RUR_URBNAME %in% c("Rural", "Urban")) 
  
ggplot(crash_filtered, aes(x = RUR_URBNAME, y = FATALS, fill = RUR_URBNAME)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values = c("Rural" = "forestgreen", "Urban" = "pink")) +
  labs(
    title = "Total Fatalities: Rural vs. Urban",
    x = "Location Type",
    y = "Total Fatalities",
    fill = "Legend"
  ) +
  theme_minimal()


ruralmodel <- aov(FATALS ~ RUR_URBNAME, data = crash_filtered)


```

Column {data-width=450 .tabset .tabset-fade}
-------------------------------------
### About

This graph shows the relationship between whether a place is classified as rural or urban, and the total pedestrian car crash fatalities in one year. As we did not have data for crashes that did not result in fatalities, this is just representative of the total number of fatalities and whether they took place in a rural or urban place.

**Research Question:** Do rural or urban places have more pedestrian fatalities? 

**Hypothesis:** I hypothesize that urban places will show more pedestrian fatalities.

**Conclusions:** Urban places showed more pedestrian fatalities than rural places. A one-way ANOVA was used to determine this relationship (*F*(33377) = 173.6, *p* < 0.001).

### first version
```{r, echo=FALSE}
ggplot(crash_filtered, aes(x = RUR_URBNAME, y = FATALS, fill = RUR_URBNAME)) +
  geom_bar(stat = "summary", fun = "sum") +
  # This line creates the separate panels for each state
  facet_wrap(~ `STATENAME`) + 
  scale_fill_manual(values = c("Rural" = "forestgreen", "Urban" = "pink")) +
  labs(
    title = "Fatalities by State: Rural vs. Urban",
    x = "Location Type",
    y = "Total Fatalities",
    fill = "Legend"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  )

```
### Rural fatalities by state 
```{r, echo=FALSE}
state_totals <- crash_filtered %>%
  group_by(STATENAME, RUR_URBNAME) %>%
  summarise(total_fatalities = sum(FATALS), .groups = "drop")

ggplot(filter(state_totals, RUR_URBNAME == "Rural"),
       aes(x = reorder(STATENAME, total_fatalities), y = total_fatalities)) +
  geom_col(fill = "forestgreen") +
  coord_flip() +
  labs(title = "Rural Fatalities by State", x = "State", y = "Total Fatalities") +
  theme_minimal()
```
### Urban fatalities by state
```{r}
ggplot(filter(state_totals, RUR_URBNAME == "Urban"),
       aes(x = reorder(STATENAME, total_fatalities), y = total_fatalities)) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(title = "Urban Fatalities by State", x = "State", y = "Total Fatalities") +
  theme_minimal()
```
### Combined plot
```{r}
state_totals <- state_totals %>%
  mutate(plot_value = ifelse(RUR_URBNAME == "Rural",
                             -total_fatalities,
                             total_fatalities))
state_totals <- state_totals %>%
  group_by(STATENAME) %>%
  mutate(total_state = sum(total_fatalities)) %>%
  ungroup()

ggplot(state_totals,
      aes(x = reorder(STATENAME, total_state), y = plot_value, fill = RUR_URBNAME)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Rural" = "forestgreen", "Urban" = "pink")) +
  scale_y_continuous(labels = abs) +
  labs(
    title = "Fatalities by State: Rural vs Urban",
    x = "State",
    y = "Total Fatalities",
    fill = "Location Type"
  ) +
  theme_minimal()

```

